---
title: "Gas Turbine CO and NOx Emission Analysis"
author: "Aayushi Gupta, Kyle Kaminski, Rosa Lin, Ruben Martinez"
date: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction 

### Gas Turbine CO and NOx Emission Data Set

The data comes from a gas turbine located in Turkey that studies the flue gas emissions of specifically carbon monoxide (CO) and nitrogen oxide (NOx) gases. The data set provides hourly statistics of 11 sensors. Data points were collected from a gas turbine from Jan 01 2011 to Dec 13 2015. 

## Description

The data file `gt_2015.csv` has 7384 observations and 11 variables from the UCI Gas Turbine CO and NOx Emission Data Set. We are going to explore and analyze the following variables: 

* `AT` - Ambient Temperature
* `AP` - Ambient Pressure
* `AH` -  Ambient Humidity
* `AFDP` - Air filter difference pressure 
* `GTEP` - Gas turbine exhaust pressure
* `TIT` - Turbine inlet temperature
* `TAT` - Turbine after temperature 
* `TEY` - Turbine energy yield 
* `CDP` - Compressor discharge pressure

Here’s a quick peek at the data set: 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
library(readr) 
gt_2015 <- read_csv("Project Data/gt_2015.csv")
knitr::kable(head(gt_2015)[,])
```

## Goals

The goal for this project is to utilize this data set for the purpose of studying flue gas emissions, specifically carbon monoxide(CO) and nitrogen oxides (NOx). Our focus will be to find statistically significant relationships between the ambient and turbine variables and the emissions variables. We will limit the size of our model to more clearly demonstrate these relationships. Ultimately we will suggest which variables make the biggest impact on emission levels in order to decrease emissions overall. 


# Methods

## Linear Regression

We will create a multiple linear regression model using all feature variables mentioned in the description of Section 1. The implementation and parameters of this model can be obtained by the following equation where we will find estimates for the parameters $$\beta$$$ using: 

$$\hat \beta = (X^TX)^{-1}X^Ty$$


Key assumptions are stated as: 

* **L**inearity: can be written as a linear combination of the predictors.
* **I**ndependence: the errors are independent of each other (not highly correlated).
* **N**ormality: the distribution of the errors follow a normal distribution.
* **E**qual Variance: the error variance is the same. (insert source, STAT 420 textbook) 


We will then use model selection using backward BIC to tune our model and remove any insignificant predictor variables. This selection prefers smaller models which aligns with our goal of limiting the size of our final model. 




